### 2025-06-08 chapters/background.tex
updated references (still in progress)

\section{Keystone architecture overview}

Keystone is an open-source Trusted Execution Environment (TEE) framework designed specifically for the RISC-V architecture. It enables the development of secure enclaves that are customizable, lightweight, and suitable for both research and deployment in resource-constrained or security-critical environments. Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features. \cite{Lee2019}

The Keystone architecture operates across the multiple privilege levels defined by the RISC-V specification: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). These privilege levels are central to the enforcement of security policies within the system. Keystone's Security Monitor (SM), the heart of its TCB, executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and enforcement of access control policies.

A typical Keystone-enabled platform consists of RISC-V processor cores augmented with secure boot mechanisms, trusted entropy sources, and optional hardware accelerators for cryptographic operations or secure I/O. Keystone does not mandate specific hardware features beyond RISC-V compliance and PMP (Physical Memory Protection), thereby supporting a broad spectrum of implementations from FPGA prototypes to silicon SoCs.

The SM, implemented in C and compiled as a standalone binary, embodies the critical logic responsible for managing the creation, execution, and destruction of enclaves. During enclave creation, the SM validates the enclave's memory layout, initializes its metadata, and configures PMP to ensure that only the SM and the enclave itself can access the designated memory regions. Once enclave execution begins, transitions between host and enclave contexts are mediated by the SM, which dynamically adjusts PMP settings to preserve strict isolation.

Each enclave is composed of two software layers: a user-level application (eapp) and a supervisor-level runtime environment. The eapp contains application-specific logic, while the runtime provides essential operating system abstractions such as exception handling, system call support, and virtual memory management. This two-tiered design allows for flexible enclave development and reduces the complexity of the eapp, which remains entirely untrusted by the host operating system.

The enclave lifecycle encompasses three primary stages: creation, execution, and destruction. During creation, the untrusted host allocates a contiguous region of physical memory—known as Enclave Page Memory (EPM)—to hold the enclave’s code, data, and page tables. This memory is populated and locked by the SM using PMP. In the execution phase, the SM facilitates entry into and exit from the enclave, ensuring that context switches maintain enclave confidentiality. Finally, destruction involves securely erasing enclave memory and releasing resources to prevent post-mortem data leakage.

Keystone also supports optional features such as remote attestation, allowing third-party verifiers to authenticate enclave state before provisioning sensitive data. Additionally, the architecture can be extended to support secure I/O pathways, cryptographic acceleration, and debugging features useful for both development and runtime monitoring.

@@

Overall, Keystone provides a clean, modular, and research-oriented TEE framework that leverages the openness of RISC-V to deliver both strong security guarantees and development flexibility. Its layered architecture, minimal TCB, and standards-compliant interfaces make it a compelling platform for building and evaluating secure applications on next-generation processors.

\section{Comparison with other TEEs}

Keystone distinguishes itself from existing commercial Trusted Execution Environment (TEE) implementations, such as Intel Software Guard Extensions (SGX) and ARM TrustZone, through its open-source nature, customizable architecture, and alignment with the RISC-V ecosystem. While SGX and TrustZone have been widely deployed in commodity hardware and offer robust isolation features, they are constrained by their proprietary designs, large and opaque TCBs, and limited extensibility, which make them less suitable for academic research and fine-grained system-level customization.

ARM TrustZone utilizes a dual-world execution model, dividing the processor into a Secure World and a Normal World. The system transitions between these worlds using secure monitor calls (SMCs). Each world operates with its own set of execution contexts and memory spaces. In the Secure World, a trusted operating system—such as OP-TEE—executes sensitive applications that conform to the GlobalPlatform TEE Internal API. However, TrustZone’s architecture supports only a single secure world instance, lacks support for dynamic enclave creation, and requires significant overhead for world switches, especially in the handling of interrupts. Moreover, the TCB in TrustZone is relatively large, encompassing the entire Secure World kernel and trusted services, which complicates formal verification and increases the risk of vulnerabilities.

Intel SGX, in contrast, supports dynamic creation of enclaves at runtime within the address space of user-space processes. These enclaves are protected by a memory encryption engine and execute in ring 3, isolated from both the operating system and hypervisor. Communication between host and enclave is facilitated by ECALL and OCALL mechanisms, whose interfaces are defined through the edger8r tool. While SGX offers strong isolation and cryptographic sealing of enclave contents, it imposes rigid constraints on enclave memory size and structure. Notably, SGX enclaves do not support supervisor-level execution, limiting the complexity of software that can be securely encapsulated within the enclave. Additionally, SGX is tightly bound to Intel hardware, restricting portability and experimentation.

Keystone integrates the strengths of both TrustZone and SGX while addressing their limitations through an open and modular framework. Its architecture supports dynamic enclave creation, hierarchical privilege separation (via user and supervisor modes), and execution control via a minimal M-mode Security Monitor. Enclave memory is isolated using RISC-V’s Physical Memory Protection (PMP) features, which are dynamically configured by the SM to enforce strict access control. This enables strong spatial and temporal memory isolation without relying on hardware-bound limitations or encryption mechanisms.

Keystone also supports a lightweight communication interface, dubbed keyedger, which serves a similar role to edger8r in SGX, but is designed to be more flexible and extensible. While it currently offers partial support for the GlobalPlatform TEE Internal API, Keystone’s modularity allows it to be extended to support custom APIs and protocols suited to specific use cases or research goals.

Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for investigating and deploying secure enclaves. It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning it as a forward-looking alternative to incumbent TEE technologies.

@@

>>>>

\section{Keystone Architecture Overview}

Keystone is an open-source Trusted Execution Environment (TEE) framework built on the RISC-V architecture, designed to enable customizable, lightweight, and secure enclaves suitable for both research and deployment in resource-constrained or security-critical settings \cite{dayeol2019keystone}. The framework’s modular design emphasizes a minimal Trusted Computing Base (TCB), portability across diverse RISC-V platforms, and extensibility to accommodate advanced security features, reflecting a clean and flexible architectural philosophy.

Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively. At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement. The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{dayeol2019keystone}.

A typical Keystone-enabled platform comprises RISC-V cores augmented with secure boot mechanisms, trusted entropy sources, and optionally, hardware accelerators for cryptographic functions or secure I/O. Importantly, Keystone does not require hardware beyond standard RISC-V compliance and PMP support, making it adaptable to implementations ranging from FPGA prototypes to silicon SoCs \cite{dayeol2019keystone}.

Each enclave in Keystone consists of two main software components: a user-level enclave application (eapp) containing application-specific logic, and a supervisor-level runtime (RT) providing essential operating system abstractions such as exception handling, system call services, and memory management. This layered approach reduces the complexity of enclave applications and provides a trusted runtime environment isolated from the untrusted operating system and other processes \cite{dayeol2019keystone}.

The enclave lifecycle is structured into three primary phases: creation, execution, and destruction. During creation, the SM validates and locks the enclave’s memory layout—referred to as Enclave Page Memory (EPM)—with PMP to ensure confidentiality and integrity. During execution, the SM mediates context switches between the host and enclave, dynamically adjusting PMP to maintain isolation. Destruction securely erases enclave memory to prevent any residual data leakage \cite{dayeol2019keystone}. Keystone also supports optional features such as remote attestation, allowing external verifiers to authenticate enclave state prior to provisioning sensitive data, and can be extended for secure I/O and cryptographic acceleration.

@@

Overall, Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization. Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}.

\section{Comparison with Other TEEs}

Keystone’s architecture sets it apart from existing commercial Trusted Execution Environments such as Intel Software Guard Extensions (SGX) and ARM TrustZone through its openness, modularity, and alignment with the RISC-V ecosystem \cite{dayeol2019keystone}. Unlike the proprietary and largely opaque designs of SGX and TrustZone, Keystone offers a fully open-source platform that supports fine-grained system-level customization and experimental research.

ARM TrustZone implements a dual-world execution model that partitions the processor into a Secure World and a Normal World, with transitions managed via secure monitor calls (SMCs). While TrustZone provides a trusted OS environment (e.g., OP-TEE) within the Secure World, it supports only a single secure instance and requires significant overhead for context switches and interrupt handling. Moreover, the TrustZone TCB encompasses the entire Secure World kernel and trusted services, which increases complexity and hinders formal verification \cite{dayeol2019keystone}.

Intel SGX allows dynamic creation of enclaves at runtime within user-mode processes and provides memory encryption to isolate enclave data from the OS and hypervisor. However, SGX enclaves are confined to user mode and cannot run supervisor-level code, limiting their expressiveness. Additionally, SGX enforces rigid constraints on enclave size and structure, and its reliance on Intel hardware restricts portability and adaptability \cite{dayeol2019keystone}.

Keystone synthesizes the advantages of both approaches while overcoming their limitations. It supports dynamic enclave creation and hierarchical privilege separation by utilizing both user and supervisor modes, controlled by a minimal M-mode Security Monitor. The use of PMP allows Keystone to enforce spatial and temporal memory isolation without the complexity of encryption or hardware-bound restrictions \cite{dayeol2019keystone}.

Communication between the host and enclaves is facilitated by a lightweight interface named keyedger, akin to SGX’s edger8r but designed for greater flexibility and extensibility. Keystone partially supports the GlobalPlatform TEE Internal API and is architected to accommodate custom APIs, enhancing its utility for diverse research and deployment scenarios \cite{dayeol2019keystone}.

Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for secure enclave development. It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning itself as a forward-looking alternative to existing TEE technologies \cite{dayeol2019keystone}.

@@

Trusted Execution Environments (TEEs) provide strong isolation guarantees for sensitive computations, but often at the cost of performance overheads. These overheads arise from architectural constraints, such as context switching, secure memory management, and the need to isolate execution from potentially untrusted system components. Evaluating these overheads is crucial for determining the feasibility of TEE deployment in latency-sensitive or resource-constrained applications.

Accurate time measurement within TEEs presents a significant challenge. Standard API functions such as \texttt{TEE\_GetREETime()} incur high latencies due to context switching into the Rich Execution Environment (REE). On the other hand, functions like \texttt{TEE\_GetSystemTime()}, which access hardware timers within the TEE, exhibit significantly lower and more consistent overhead. In comparative tests, \texttt{TEE\_GetSystemTime()} was approximately 30 times faster on average than \texttt{TEE\_GetREETime()} across various architectures including ARM TrustZone, Intel SGX, and RISC-V Keystone \cite{Suzaki2021}.

Benchmark results further confirm that CPU-intensive workloads experience minimal overhead within TEEs, as these operations typically rely on core-local instruction pipelines. For example, integer and floating-point multiplications executed within a TEE yielded results comparable to their REE counterparts across platforms, with ARM TrustZone showing a modest 3–4\% slowdown \cite{Suzaki2021}. These results were consistent with Keystone benchmarks using RV8 and CoreMark, which reported sub-1\% overhead for CPU-bound tasks when enclave lifecycle costs were excluded \cite{dayeol2019keystone}.

In contrast, memory and I/O operations often incur greater penalties. In TS-perf benchmarks, ARM TrustZone showed an 11\% performance drop in random memory access, attributed to memory encryption and architectural characteristics such as cache hierarchy. Further experiments revealed that performance degradation in Intel SGX becomes significant only after memory footprints exceed the L2 cache size, highlighting cache capacity as a mitigating factor \cite{Suzaki2021}.

Keystone's RV8 benchmarks show that memory-heavy workloads with large working sets (e.g., \texttt{primes}, \texttt{miniz}, \texttt{aes}) can experience overheads exceeding 100\% when cache partitioning is enabled, due to reduced effective cache capacity and increased eviction rates. Conversely, applications with small memory footprints (e.g., \texttt{sha512}, \texttt{dhrystone}) demonstrated negligible impact from cache-related protections \cite{dayeol2019keystone}. This suggests that performance optimization in TEEs requires close alignment between application memory behavior and enclave hardware configuration.

Storage performance is another critical consideration. TEE implementations typically must tunnel file I/O through the REE, using secure APIs that copy data into untrusted buffers and invoke enclave exit calls. This indirection introduces substantial overhead. In Keystone, file operations using IOZone showed throughput reductions of 36.2\% (write) and 40.9\% (read) on average, with even higher penalties on record re-access patterns (e.g., 55.1\% on record rewrite) \cite{dayeol2019keystone}. TS-perf results confirm similar patterns, with OP-TEE on ARM and SGX showing large performance disparities between REE and TEE storage operations due to the reliance on OCALLs and limited TEE-side file APIs \cite{Suzaki2021}.

Furthermore, platform-specific runtime behavior can introduce variability in performance. For example, htop monitoring during TS-perf experiments revealed unexpected behavior in thread scheduling: TEE applications sometimes migrated between CPU cores or were scheduled differently from their REE parent processes. This inconsistency suggests a lack of TEE-awareness in existing OS schedulers, particularly in OP-TEE and Keystone. In some cases, CPU load dropped to zero despite active enclave execution, later traced to interrupt handling bugs in RISC-V’s eyrie runtime \cite{Suzaki2021}.

Finally, enclave lifecycle operations, while one-time costs, also contribute to overall performance. In Keystone, enclave creation and destruction involve validation, hashing (e.g., SHA-3), and context setup, incurring 2M–7M cycles per page for validation and 20k–30k cycles for lifecycle events. These costs are significantly affected by software-based implementations of cryptographic primitives and could be optimized using hardware acceleration \cite{dayeol2019keystone}.

In summary, while TEEs offer minimal performance penalties for compute-bound workloads, substantial overheads can emerge for memory- and I/O-bound applications. These overheads are highly platform-dependent and are influenced by factors such as cache behavior, buffer copying, interrupt handling, and runtime design. Hence, effective use of TEEs requires not only security-aware application design but also performance-aware system integration.
